Ship-based lidar evaluation of Southern Ocean clouds in the storm-resolving general circulation model ICON, and the ERA5 and MERRA-2 reanalyses

### 2024-12-12 manuscript_agu.tex
Indivisible space

\begin{keypoints}
\item Ground-based lidar evaluation of a km-scale global climate model and reanalyses reveals substantial cloud biases over the Southern Ocean.
\item Fog or low cloud is underestimated in the reanalyses. In all models, a cloud peak at 500 m tends to be overestimated and too high.
\item A ``too few, too bright'' problem of underestimated cloud fraction, compensated by overestimated cloud albedo, is present in the reanalyses.
\end{keypoints}

\begin{abstract}
Global storm-resolving models (GSRMs) are the upcoming global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). The high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but still utilizing microphysics and turbulence parameterizations. Standard-resolution (10--100 km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using about 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and a station during 2010--2021, compared with the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500 m, potentially explained by their lifting condensation levels being too high. The reanalyses strongly underestimate fog or near-surface clouds, and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in the reanalyses, implying a ``too few, too bright'' cloud problem. Thermodynamic conditions are relatively well represented, but ICON is less stable than observations and MERRA-2 is too humid. SO cloud biases are a substantial issue in the GSRM, but it matches the observations better than the lower-resolution reanalyses.
\end{abstract}

@@

Increasing climate model resolution is one way of improving model accuracy of representation of the climate system \cite{mauritsen2022}. It has been practiced since the advent of climate modeling as more computational power, memory, and storage capacity become available. It is, however, often not as easy as changing the grid size because of the complex interplay between model dynamics and physics, which necessitates adjusting and tuning all components together. Increasing resolution is, of course, limited by the available computational power and a trade-off with increasing parameterization complexity, which is another way of improving model accuracy. Current computational availability and acceleration from general-purpose computing on graphics processing units (GPUs) is progressing to enable km-scale (also called k-scale) Earth system models (ESMs) and coupled atmosphere--ocean general circulation models (AOGCMs) in research conditions today and operationally in the forthcoming years. Therefore, it represents a natural advance in climate modeling. Global storm-resolving models (GSRMs) are emerging as a new front in the development of high-resolution global climate models, with horizontal grid resolutions of about 2--8 km \cite{satoh2019,stevens2019}. This is enough to resolve mesoscale convective storms, but smaller-scale convective plumes and cloud structure remain unresolved. At an approximately 5-km scale, non-hydrostatic processes also become important \cite{weisman1997}, and for this reason such models are generally non-hydrostatic. The terms global cloud-resolving models or global convection-permitting/-resolving models are also sometimes used interchangeably with GSRMs but imply that clouds or convection are resolved explicitly, which is not entirely true for GSRMs, as this would require an even higher horizontal resolution \cite{satoh2019}. Representative of these efforts is the DYnamics of the Atmospheric general circulation Modeled On Non-hydrostatic Domains (DYAMOND) project \cite{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs over two 40-day time periods in summer (1 August -- 10 September 2016) and winter (20 January -- 1 March 2020). A new one-year GSRM intercomparison is currently proposed by \citeA{takasuka2024}, with the hope of also evaluating the seasonal cycle and large-scale circulation. An alternative to using a computationally costly GSRM is to train an artificial neural network on GSRM output and use it for subgrid-scale clouds, as done with the GSRM ICON by \citeA{grundner2022} and \citeA{grundner2023}.

@@

Multiple cloud properties have an effect on shortwave (SW) and longwave (LW) radiation. To first order, the total cloud fraction, cloud phase, and the liquid and ice water path are the most important cloud properties influencing SW and LW radiation. These properties are in turn influenced by the atmospheric thermodynamics, convection and circulation, and both the indirect and direct effects of aerosols. Second-order effects on SW and LW radiation are associated with the cloud droplet size distribution, ice crystal habit, cloud lifetime, and direct radiative interaction with aerosols. In the 6$^\mathrm{th}$ phase of the Coupled Model Intercomparison Project [CMIP6; \citeA{eyring2016}], the cloud feedback has increased relative to CMIP5 \cite{zelinka2020}, which is one of the main reasons for the higher climate sensitivity of CMIP6 models.

@@

SO radiation biases have been relatively large and systematic compared to the rest of the globe since at least CMIP3 \cite{trenberth2010}, and the SO SW cloud radiative effect (CRE) bias is still positive in eight analyzed CMIP6 models analyzed by \citeA{schuddeboom2021} over the high-latitude SO, whereas over the low-latitude SO it tends to be more neutral or negative in some models. Too much absorbed SW radiation over the SO was also identified in the GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the ``too few too bright'' cloud bias, characterized by too small cloud fraction and too large cloud albedo \cite{wall2017,kuma2020}, previously described by \citeA{webb2001}, \citeA{weare2004}, \citeA{zhang2005}, \citeA{karlsson2008}, \citeA{nam2012}, \citeA{klein2013}, and \citeA{bender2017} in other regions and models, which means that a model can maintain a reasonable SW radiation balance by reflecting too much SW radiation from clouds, but these cover too small an area. A study by \citeA{konsta2022} showed that this type of bias is still present in six analyzed CMIP6 models in tropical marine clouds, using the GCM-Oriented Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO) Cloud Product [CALIPSO--GOCCP; \citeA{chepfer2010}] and Polarization \& Anisotropy of Reflectances for Atmospheric Sciences coupled with Observations from a Lidar [PARASOL; \citeA{lier2008}] as a reference. They suggest improper simulation of subgrid-scale cloud heterogeneity as a cause. Compensating cloud biases in the Australian Community Climate and Earth System Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were analyzed by \citeA{fiddes2022} and \citeA{fiddes2024}. \citeA{possner2022} showed that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction on the order of 30\% and overestimates net downward TOA SW radiation by approximately 10 Wm$^\mathrm{-2}$ in the highest model resolution run (2.5 km). \citeA{zhao2022} reported a similar SW radiation bias in five analyzed CMIP6 models over high-latitude SO and the total cloud fraction underestimation on the order of 10\% over the entire 40--60°S SO. Recently, \citeA{ramadoss2024} analyzed 48 hours of km-scale ICON limited-area model NWP simulations over a SO region adjacent to Tasmania against the Clouds, Aerosols, Precipitation, Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN) voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found the ICON cloud optical thickness was underestimated relative to Himawari‐8 satellite observations but also identified large differences in cloud top phase.

@@

Altogether, we analyzed about 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the model, we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with ERA5 \cite{era5} and the Modern-Era Retrospective analysis for Research and Applications, Version 2 [MERRA-2; \citeA{gelaro2017}].

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/map_fig.pdf}
\caption{
\textbf{(a)} A map showing the tracks of 31 voyages of RV \emph{Polarstern}, RSV \emph{Aurora Australis}, RV \emph{Tangaroa}, RV \emph{Nathaniel B. Palmer}, and HMNZS \emph{Wellington} and one sub-Antarctic station (Macquarie Island) analyzed here. The tracks cover Antarctic sectors south of South America, the Atlantic Ocean, Africa, Australia, and New Zealand in the years 2010--2021 (inclusive). The dotted and dashed lines at 40°S and 55°S delineate the Southern Ocean area of our analysis and its partitioning into two subsets, respectively. A photo of \textbf{(b)} RV \emph{Polarstern} (© Folke Mehrtens, Alfred-Wegener-Institut), \textbf{(c)} Lufft CHM 15k installed on RV \emph{Tangaroa} (© Peter Kuma, University of Canterbury), \textbf{(d)} Vaisala CL51 (© Jeff Aquilina, Bureau of Meteorology), \textbf{(e)} Vaisala CT25K at Macquarie Island (© Simon P.\ Alexander, Australian Antarctic Division).
}
\label{fig:map}
\end{figure}

@@

Together, we analyzed data from 31 voyages of RV \emph{Polarstern}, the resupply vessel (RSV) \emph{Aurora Australis}, RV \emph{Tangaroa}, RV \emph{Nathaniel B. Palmer}, Her (now His) Majesty's New Zealand Ship (HMNZS) \emph{Wellington} and one sub-Antarctic station (Macquarie Island) in the SO south of 40°S between 2010 and 2021. Fig.\ \ref{fig:map} shows a map of the campaigns, Table \ref{tab:voyages} lists the campaigns, and Table \ref{tab:voyage-references} lists references where available. Altogether, the analyzed dataset comprised 2421 days of data south of 40°S, but the availability of ceilometer data was slightly shorter due to gaps in measurements.

The campaigns contained ceilometer observations captured by the Vaisala CL51 and CT25K, and the Lufft CHM 15k, described in detail below (Sections\ \ref{sec:cl51} and \ref{sec:chm15k}). A ceilometer is a low-power, near-infrared, vertically pointing lidar principally designed to measure cloud base, but they also measure the full vertical structure of clouds as long as the laser signal is not attenuated by thick clouds, which can be used to infer additional information such as a cloud mask and cloud occurrence by height. We note that during the MICRE campaign, the ceilometers Vaisala CT25K and CL51 were installed at the Macquarie Island station concurrently, but in our analysis we only used the CT25K data obtained from the Atmospheric Radiation Measurement (ARM) data archive.

@@

An overview of the analyzed campaigns (voyages and stations). Start, end, and the number of days (UTC; inclusive) refer to the time period when the vessel was south of 40°S. Abbreviations: ceilometer (ceil.), Australia (AU), New Zealand (NZ), South America (SA), Atlantic Ocean (AO), and Africa (AF). The number of days is rounded to the nearest integer. CL51/31 indicates CL51 configured to emulate CL31. Missing days in the ceilometer data were HMNZSW16 (7 days): 24--27 November, 10 December, 16--17 December 2016; MARCUS (3 days): 8, 10 November, 10 December 2017; MICRE (9 days): 7--8, 29 June, 5, 16 July, 15 August, 17 October 2016, 11 February, 21 March 2017; TAN1502 (1 day): 24 January.

@@

The Vaisala CL51 and CT25K (photos in Fig.\ \ref{fig:map}d, e) are ceilometers operating at near-infrared wavelengths of 910 nm and 905 nm, respectively. The CL51 can also be configured to emulate the Vaisala CL31. The maximum range is 15.4 km (CL51), 7.7 km (CL31 emulation mode with 5 m vertical resolution), and 7.5 km (CT25K). The vertical resolution is 10 m (5 m configurable) in CL51 and 30 m in CT25K observations. The sampling (temporal) resolution is configurable, and in our datasets, it is approximately 6 s for CL51 on AA15‐16, 16 s for CT25K on MARCUS and MICRE, 36 s for CL51 on RV \emph{Polarstern}, and about 2.37 s for CL51 with CL31 emulation on TAN1502. The wavelengths of 905 and 910 nm are affected by water vapor absorption of about 20\% in the mid-latitudes \cite{wiegner2015,wiegner2019}, with 910 nm affected more strongly, but we do not expect this to be a significant issue as explained in \citeA{kuma2021}. The instrument data files containing raw uncalibrated backscatter were first converted to Network Common Data Form (NetCDF) with cl2nc (\url{https://github.com/peterkuma/cl2nc}) and then processed with the ALCF (Section\ \ref{sec:alcf}) to produce absolutely calibrated attenuated volume backscattering coefficient (AVBC), cloud mask, cloud occurrence by height, and the total cloud fraction. Because the CT25K uses a very similar wavelength to CL51, equivalent calculations as for CL51 were done assuming a wavelength of 910 nm. The Vaisala CL51 and CT25K instruments were used on most of the voyages and stations analyzed here. Fig.\ \ref{fig:example}a shows an example of AVBC derived from the CL51 instrument data.

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/example.png}
\caption{
An example of the attenuated volume backscattering coefficient (AVBC) \textbf{(a)} measured by CL51 during 24 hours on the PS81/3 voyage and \textbf{(b)} an equivalent AVBC simulated with the ALCF from ERA5 data during the same time period. The red line identifies the cloud mask determined by the ALCF.
}
\label{fig:example}
\end{figure}

\subsection{Lufft CHM 15k}
\label{sec:chm15k}

The Lufft CHM 15k (photo in Fig.\ \ref{fig:map}c) is a ceilometer operating at a near-infrared wavelength of 1064 nm. The maximum range is 15.4 km; the vertical resolution is 5 m in the near range (up to 150 m) and 15 m above; the sampling (temporal) resolution is 2 s; and the number of vertical levels is 1024. NetCDF files containing uncalibrated backscatter produced by the instrument were processed with the ALCF (Section\ \ref{sec:alcf}) to again produce AVBC, cloud mask, cloud occurrence by height, and the total cloud fraction. The CHM 15k was used on four voyages (HMNZSW16, TAN1702, TAN1802, and NBP1704).

@@

Absolute calibration of the observed backscatter was performed by comparing the measured clear-sky molecular backscatter statistically with simulated clear-sky molecular backscatter. AVBC was resampled to 5 min temporal resolution and 50 m vertical resolution to increase the signal-to-noise ratio while having enough resolution to detect small-scale cloud variability. The noise standard deviation was calculated from AVBC at the highest range, where no clouds are expected. A cloud mask was calculated from AVBC using a fixed threshold of $\mathrm{2\times 10^{-6} m^{-1}sr^{-1}}$ after subtracting 5 standard deviations of range-scaled noise. Fig.\ \ref{fig:example}b shows an example of simulated Vaisala CL51 backscatter from ERA5 data, corresponding to a day of measurements by the instrument on the PS81/3 voyage.

@@

Unlike current general circulation models (GCMs), the storm-resolving version of ICON does not use convective and cloud parameterization but relies on explicit simulation of convection and clouds on the model grid. Subgrid-scale clouds are not resolved, and the grid cell cloud fraction is always either 0 or 100\%. While this makes the code development simpler without having to rely on uncertain parameterizations, it can miss smaller-scale clouds below the grid resolution. Turbulence and cloud microphysics are still parameterized in this model, and aerosols are taken from a climatology. To account for the radiative effects of subgrid-scale clouds, a cloud inhomogeneity factor is introduced in the model, which scales down the cloud liquid water for radiative calculations. It ranges from 0.4 at lower tropospheric stability (LTS) of 0 K to 0.8 at 30 K. In addition, vertical mixing is enhanced in unstable and humid lower-tropospheric conditions, which reduces the amount of shallow clouds.

@@

ERA5 \cite{era5} is a reanalysis produced by the ECMWF. It is based on a numerical weather prediction model IFS version CY41R2. It uses the \citeA{tiedtke1993} prognostic cloud scheme and \citeA{forbes2014} for mixed-phase clouds. The horizontal resolution is 0.25° in latitude and longitude, which is about 28 km in the North--South direction and 14 km in the East--West direction at 60°S. Internally, the model uses 137 vertical levels. Here, we use output at 1-hourly instantaneous time intervals, except for radiation quantities, which are accumulations (from these we calculate daily means). Vertically resolved quantities are made available on 37 pressure levels.

@@

On the RV \emph{Polarstern} voyages, regular human-performed synoptic observations were available and included precipitation presence or absence and type. We used this dataset to train a convolutional artificial neural network (ANN) to recognize profiles with precipitation from lidar backscatter (Fig.\ \ref{fig:ann}a), implemented in the TensorFlow ANN framework \cite{tensorflow}. Samples of short time intervals (10 min) of near-surface lidar backscatter (0–250 m) were classified as clear, rain, snow, and fog, using the synoptic observations as a training dataset (Fig.\ \ref{fig:ann}b). From these, a binary, mutually exclusive classification of profiles as precipitating (rain or snow) or dry (clear or fog) was derived. For detecting model and reanalysis precipitation, we used a fixed threshold for surface precipitation flux of 0.1 mm h$^{-1}$ (the ANN was not used).

@@

In addition to the above, we partitioned our data into two mutually exclusive subsets based on LTS, which is derived as the difference between the potential temperature at 700 hPa and the surface. Based on a histogram of LTS in ERA5 and MERRA-2 calculated at all voyage tracks and stations (Fig.\ \ref{fig:lts}), we determined a statistically-based dividing threshold of 12 K for weak stability ($<$ 12 K) and strong stability ($\geq$ 12 K) conditions.

@@

Ground-based lidar observations are affected by attenuation by thick cloud layers, and for this reason the results are most representative of boundary layer clouds, while higher-level clouds are only occasionally visible to the lidar when boundary layer clouds are not present. Ground-based lidar observations can be regarded as superior to satellite lidar observations for low-level clouds, which are predominant in this region, while mid- and high-level clouds are better sampled by satellite observations \cite{mcerlich2021}. Near-surface lidar retrievals (~100 m) are affected by uncertainties related to incomplete overlap, signal saturation (dead time), and after-pulse effect corrections \cite{kuma2021}.

>>>>

\begin{keypoints}
\item Ground-based lidar evaluation of a km-scale global climate model and reanalyses reveals substantial cloud biases over the Southern Ocean.
\item Fog or low cloud is underestimated in the reanalyses. In all models, a cloud peak at 500\ m tends to be overestimated and too high.
\item A ``too few, too bright'' problem of underestimated cloud fraction, compensated by overestimated cloud albedo, is present in the reanalyses.
\end{keypoints}

\begin{abstract}
Global storm-resolving models (GSRMs) are the upcoming global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). The high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but still utilizing microphysics and turbulence parameterizations. Standard-resolution (10--100\ km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using about 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and a station during 2010--2021, compared with the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500 m, potentially explained by their lifting condensation levels being too high. The reanalyses strongly underestimate fog or near-surface clouds, and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in the reanalyses, implying a ``too few, too bright'' cloud problem. Thermodynamic conditions are relatively well represented, but ICON is less stable than observations and MERRA-2 is too humid. SO cloud biases are a substantial issue in the GSRM, but it matches the observations better than the lower-resolution reanalyses.
\end{abstract}

@@

Increasing climate model resolution is one way of improving model accuracy of representation of the climate system \cite{mauritsen2022}. It has been practiced since the advent of climate modeling as more computational power, memory, and storage capacity become available. It is, however, often not as easy as changing the grid size because of the complex interplay between model dynamics and physics, which necessitates adjusting and tuning all components together. Increasing resolution is, of course, limited by the available computational power and a trade-off with increasing parameterization complexity, which is another way of improving model accuracy. Current computational availability and acceleration from general-purpose computing on graphics processing units (GPUs) is progressing to enable km-scale (also called k-scale) Earth system models (ESMs) and coupled atmosphere--ocean general circulation models (AOGCMs) in research conditions today and operationally in the forthcoming years. Therefore, it represents a natural advance in climate modeling. Global storm-resolving models (GSRMs) are emerging as a new front in the development of high-resolution global climate models, with horizontal grid resolutions of about 2--8\ km \cite{satoh2019,stevens2019}. This is enough to resolve mesoscale convective storms, but smaller-scale convective plumes and cloud structure remain unresolved. At an approximately 5-km scale, non-hydrostatic processes also become important \cite{weisman1997}, and for this reason such models are generally non-hydrostatic. The terms global cloud-resolving models or global convection-permitting/-resolving models are also sometimes used interchangeably with GSRMs but imply that clouds or convection are resolved explicitly, which is not entirely true for GSRMs, as this would require an even higher horizontal resolution \cite{satoh2019}. Representative of these efforts is the DYnamics of the Atmospheric general circulation Modeled On Non-hydrostatic Domains (DYAMOND) project \cite{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs over two 40-day time periods in summer (1\ August--10\ September 2016) and winter (20\ January--1\ March 2020). A new one-year GSRM intercomparison is currently proposed by \citeA{takasuka2024}, with the hope of also evaluating the seasonal cycle and large-scale circulation. An alternative to using a computationally costly GSRM is to train an artificial neural network on GSRM output and use it for subgrid-scale clouds, as done with the GSRM ICON by \citeA{grundner2022} and \citeA{grundner2023}.

@@

Multiple cloud properties have an effect on shortwave (SW) and longwave (LW) radiation. To first order, the total cloud fraction, cloud phase, and the liquid and ice water path are the most important cloud properties influencing SW and LW radiation. These properties are in turn influenced by the atmospheric thermodynamics, convection and circulation, and both the indirect and direct effects of aerosols. Second-order effects on SW and LW radiation are associated with the cloud droplet size distribution, ice crystal habit, cloud lifetime, and direct radiative interaction with aerosols. In the 6$^\mathrm{th}$\ phase of the Coupled Model Intercomparison Project [CMIP6; \citeA{eyring2016}], the cloud feedback has increased relative to CMIP5 \cite{zelinka2020}, which is one of the main reasons for the higher climate sensitivity of CMIP6 models.

@@

SO radiation biases have been relatively large and systematic compared to the rest of the globe since at least CMIP3 \cite{trenberth2010}, and the SO SW cloud radiative effect (CRE) bias is still positive in eight analyzed CMIP6 models analyzed by \citeA{schuddeboom2021} over the high-latitude SO, whereas over the low-latitude SO it tends to be more neutral or negative in some models. Too much absorbed SW radiation over the SO was also identified in the GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the ``too few too bright'' cloud bias, characterized by too small cloud fraction and too large cloud albedo \cite{wall2017,kuma2020}, previously described by \citeA{webb2001}, \citeA{weare2004}, \citeA{zhang2005}, \citeA{karlsson2008}, \citeA{nam2012}, \citeA{klein2013}, and \citeA{bender2017} in other regions and models, which means that a model can maintain a reasonable SW radiation balance by reflecting too much SW radiation from clouds, but these cover too small an area. A study by \citeA{konsta2022} showed that this type of bias is still present in six analyzed CMIP6 models in tropical marine clouds, using the GCM-Oriented Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO) Cloud Product [CALIPSO--GOCCP; \citeA{chepfer2010}] and Polarization \& Anisotropy of Reflectances for Atmospheric Sciences coupled with Observations from a Lidar [PARASOL; \citeA{lier2008}] as a reference. They suggest improper simulation of subgrid-scale cloud heterogeneity as a cause. Compensating cloud biases in the Australian Community Climate and Earth System Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were analyzed by \citeA{fiddes2022} and \citeA{fiddes2024}. \citeA{possner2022} showed that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction on the order of 30\% and overestimates net downward TOA SW radiation by approximately 10 Wm$^\mathrm{-2}$ in the highest model resolution run (2.5\ km). \citeA{zhao2022} reported a similar SW radiation bias in five analyzed CMIP6 models over high-latitude SO and the total cloud fraction underestimation on the order of 10\% over the entire 40--60°S SO. Recently, \citeA{ramadoss2024} analyzed 48\ hours of km-scale ICON limited-area model NWP simulations over a SO region adjacent to Tasmania against the Clouds, Aerosols, Precipitation, Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN) voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found the ICON cloud optical thickness was underestimated relative to Himawari‐8 satellite observations but also identified large differences in cloud top phase.

@@

Altogether, we analyzed about 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the model, we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with ERA5 \cite{era5} and the Modern-Era Retrospective analysis for Research and Applications, Version\ 2 [MERRA-2; \citeA{gelaro2017}].

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/map_fig.pdf}
\caption{
\textbf{(a)} A map showing the tracks of 31\ voyages of RV \emph{Polarstern}, RSV \emph{Aurora Australis}, RV \emph{Tangaroa}, RV \emph{Nathaniel B. Palmer}, and HMNZS \emph{Wellington} and one sub-Antarctic station (Macquarie Island) analyzed here. The tracks cover Antarctic sectors south of South America, the Atlantic Ocean, Africa, Australia, and New Zealand in the years 2010--2021 (inclusive). The dotted and dashed lines at 40°S and 55°S delineate the Southern Ocean area of our analysis and its partitioning into two subsets, respectively. A photo of \textbf{(b)} RV \emph{Polarstern} (© Folke Mehrtens, Alfred-Wegener-Institut), \textbf{(c)} Lufft CHM\ 15k installed on RV \emph{Tangaroa} (© Peter Kuma, University of Canterbury), \textbf{(d)} Vaisala CL51 (© Jeff Aquilina, Bureau of Meteorology), \textbf{(e)} Vaisala CT25K at Macquarie Island (© Simon P.\ Alexander, Australian Antarctic Division).
}
\label{fig:map}
\end{figure}

@@

Together, we analyzed data from 31\ voyages of RV \emph{Polarstern}, the resupply vessel (RSV) \emph{Aurora Australis}, RV \emph{Tangaroa}, RV \emph{Nathaniel B. Palmer}, Her (now His) Majesty's New Zealand Ship (HMNZS) \emph{Wellington} and one sub-Antarctic station (Macquarie Island) in the SO south of 40°S between 2010 and 2021. Fig.\ \ref{fig:map} shows a map of the campaigns, Table \ref{tab:voyages} lists the campaigns, and Table \ref{tab:voyage-references} lists references where available. Altogether, the analyzed dataset comprised 2421\ days of data south of 40°S, but the availability of ceilometer data was slightly shorter due to gaps in measurements.

The campaigns contained ceilometer observations captured by the Vaisala CL51 and CT25K, and the Lufft CHM\ 15k, described in detail below (Sections\ \ref{sec:cl51} and \ref{sec:chm15k}). A ceilometer is a low-power, near-infrared, vertically pointing lidar principally designed to measure cloud base, but they also measure the full vertical structure of clouds as long as the laser signal is not attenuated by thick clouds, which can be used to infer additional information such as a cloud mask and cloud occurrence by height. We note that during the MICRE campaign, the ceilometers Vaisala CT25K and CL51 were installed at the Macquarie Island station concurrently, but in our analysis we only used the CT25K data obtained from the Atmospheric Radiation Measurement (ARM) data archive.

@@

An overview of the analyzed campaigns (voyages and stations). Start, end, and the number of days (UTC; inclusive) refer to the time period when the vessel was south of 40°S. Abbreviations: ceilometer (ceil.), Australia (AU), New Zealand (NZ), South America (SA), Atlantic Ocean (AO), and Africa (AF). The number of days is rounded to the nearest integer. CL51/31 indicates CL51 configured to emulate CL31. Missing days in the ceilometer data were HMNZSW16 (7 days): 24--27 November, 10\ December, 16--17\ December 2016; MARCUS (3 days): 8, 10\ November, 10\ December 2017; MICRE (9\ days): 7--8, 29\ June, 5, 16\ July, 15\ August, 17\ October 2016, 11\ February, 21\ March 2017; TAN1502 (1\ day): 24\ January.

@@

The Vaisala CL51 and CT25K (photos in Fig.\ \ref{fig:map}d, e) are ceilometers operating at near-infrared wavelengths of 910\ nm and 905\ nm, respectively. The CL51 can also be configured to emulate the Vaisala CL31. The maximum range is 15.4\ km (CL51), 7.7\ km (CL31 emulation mode with 5\ m vertical resolution), and 7.5\ km (CT25K). The vertical resolution is 10\ m (5\ m configurable) in CL51 and 30\ m in CT25K observations. The sampling (temporal) resolution is configurable, and in our datasets, it is approximately 6\ s for CL51 on AA15‐16, 16\ s for CT25K on MARCUS and MICRE, 36\ s for CL51 on RV \emph{Polarstern}, and about 2.37\ s for CL51 with CL31 emulation on TAN1502. The wavelengths of 905 and 910\ nm are affected by water vapor absorption of about 20\% in the mid-latitudes \cite{wiegner2015,wiegner2019}, with 910\ nm affected more strongly, but we do not expect this to be a significant issue as explained in \citeA{kuma2021}. The instrument data files containing raw uncalibrated backscatter were first converted to Network Common Data Form (NetCDF) with cl2nc (\url{https://github.com/peterkuma/cl2nc}) and then processed with the ALCF (Section\ \ref{sec:alcf}) to produce absolutely calibrated attenuated volume backscattering coefficient (AVBC), cloud mask, cloud occurrence by height, and the total cloud fraction. Because the CT25K uses a very similar wavelength to CL51, equivalent calculations as for CL51 were done assuming a wavelength of 910\ nm. The Vaisala CL51 and CT25K instruments were used on most of the voyages and stations analyzed here. Fig.\ \ref{fig:example}a shows an example of AVBC derived from the CL51 instrument data.

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/example.png}
\caption{
An example of the attenuated volume backscattering coefficient (AVBC) \textbf{(a)} measured by CL51 during 24\ hours on the PS81/3 voyage and \textbf{(b)} an equivalent AVBC simulated with the ALCF from ERA5 data during the same time period. The red line identifies the cloud mask determined by the ALCF.
}
\label{fig:example}
\end{figure}

\subsection{Lufft CHM\ 15k}
\label{sec:chm15k}

The Lufft CHM\ 15k (photo in Fig.\ \ref{fig:map}c) is a ceilometer operating at a near-infrared wavelength of 1064\ nm. The maximum range is 15.4\ km; the vertical resolution is 5\ m in the near range (up to 150\ m) and 15\ m above; the sampling (temporal) resolution is 2\ s; and the number of vertical levels is 1024. NetCDF files containing uncalibrated backscatter produced by the instrument were processed with the ALCF (Section\ \ref{sec:alcf}) to again produce AVBC, cloud mask, cloud occurrence by height, and the total cloud fraction. The CHM\ 15k was used on four voyages (HMNZSW16, TAN1702, TAN1802, and NBP1704).

@@

Absolute calibration of the observed backscatter was performed by comparing the measured clear-sky molecular backscatter statistically with simulated clear-sky molecular backscatter. AVBC was resampled to 5 min temporal resolution and 50\ m vertical resolution to increase the signal-to-noise ratio while having enough resolution to detect small-scale cloud variability. The noise standard deviation was calculated from AVBC at the highest range, where no clouds are expected. A cloud mask was calculated from AVBC using a fixed threshold of $\mathrm{2\times 10^{-6} m^{-1}sr^{-1}}$ after subtracting 5 standard deviations of range-scaled noise. Fig.\ \ref{fig:example}b shows an example of simulated Vaisala CL51 backscatter from ERA5 data, corresponding to a day of measurements by the instrument on the PS81/3 voyage.

@@

Unlike current general circulation models (GCMs), the storm-resolving version of ICON does not use convective and cloud parameterization but relies on explicit simulation of convection and clouds on the model grid. Subgrid-scale clouds are not resolved, and the grid cell cloud fraction is always either 0 or 100\%. While this makes the code development simpler without having to rely on uncertain parameterizations, it can miss smaller-scale clouds below the grid resolution. Turbulence and cloud microphysics are still parameterized in this model, and aerosols are taken from a climatology. To account for the radiative effects of subgrid-scale clouds, a cloud inhomogeneity factor is introduced in the model, which scales down the cloud liquid water for radiative calculations. It ranges from 0.4 at lower tropospheric stability (LTS) of 0\ K to 0.8 at 30\ K. In addition, vertical mixing is enhanced in unstable and humid lower-tropospheric conditions, which reduces the amount of shallow clouds.

@@

ERA5 \cite{era5} is a reanalysis produced by the ECMWF. It is based on a numerical weather prediction model IFS version CY41R2. It uses the \citeA{tiedtke1993} prognostic cloud scheme and \citeA{forbes2014} for mixed-phase clouds. The horizontal resolution is 0.25° in latitude and longitude, which is about 28\ km in the North--South direction and 14\ km in the East--West direction at 60°S. Internally, the model uses 137 vertical levels. Here, we use output at 1-hourly instantaneous time intervals, except for radiation quantities, which are accumulations (from these we calculate daily means). Vertically resolved quantities are made available on 37 pressure levels.

@@

On the RV \emph{Polarstern} voyages, regular human-performed synoptic observations were available and included precipitation presence or absence and type. We used this dataset to train a convolutional artificial neural network (ANN) to recognize profiles with precipitation from lidar backscatter (Fig.\ \ref{fig:ann}a), implemented in the TensorFlow ANN framework \cite{tensorflow}. Samples of short time intervals (10\ min) of near-surface lidar backscatter (0–250 m) were classified as clear, rain, snow, and fog, using the synoptic observations as a training dataset (Fig.\ \ref{fig:ann}b). From these, a binary, mutually exclusive classification of profiles as precipitating (rain or snow) or dry (clear or fog) was derived. For detecting model and reanalysis precipitation, we used a fixed threshold for surface precipitation flux of 0.1 mm h$^{-1}$ (the ANN was not used).

@@

In addition to the above, we partitioned our data into two mutually exclusive subsets based on LTS, which is derived as the difference between the potential temperature at 700\ hPa and the surface. Based on a histogram of LTS in ERA5 and MERRA-2 calculated at all voyage tracks and stations (Fig.\ \ref{fig:lts}), we determined a statistically-based dividing threshold of 12\ K for weak stability ($<$\ 12\ K) and strong stability ($\geq$\ 12\ K) conditions.

@@

Ground-based lidar observations are affected by attenuation by thick cloud layers, and for this reason the results are most representative of boundary layer clouds, while higher-level clouds are only occasionally visible to the lidar when boundary layer clouds are not present. Ground-based lidar observations can be regarded as superior to satellite lidar observations for low-level clouds, which are predominant in this region, while mid- and high-level clouds are better sampled by satellite observations \cite{mcerlich2021}. Near-surface lidar retrievals (~100\ m) are affected by uncertainties related to incomplete overlap, signal saturation (dead time), and after-pulse effect corrections \cite{kuma2021}.
